A linear inside-outside algorithm for correcting sequencing errors in structured RNAs

### 2012-10-07 Recomb/results_RECOMB.tex
new figs
results section completed

We build our data set from the 5S rRNA multiple sequence alignment (MSA) available in the Rfam Database 11.0 (Rfam id: \texttt{RF00001}).
Since our software does not currently implement gaps (mainly because scoring indels is a challenging issue that cannot be fully addressed
in this work),  we clustered together the sequences with identical gap locations. From the $54$ MSAs without gap produced, we selected the
biggest MSA  which contains $130$ sequences (out of $712$ in the original Rfam MSA). Then, in order to avoid any bias, we used \texttt{cd-hit}
\cite{CDHIT} to remove sequences with more than 80\% of sequence similarity. This operation resulted in a data set of $45$ sequences.

We design our benchmark using a leave-one-out strategy. We randomly picked one sequence from our data set and performed $12$ random
mutations. Our sequences have $119$ nucleotides, thus the number of mutations corresponds to an error-rate of 10\%. We repeated this operation 
$10$ times.

To evaluate our method, we computed a ROC curve representing the performance of a classifier based on the mutational probabilities computed by
\RNApyro. We report in Table \ref{tab:benchmark} the area under the curve (AUC). More specifically, we fix a threshold $\lambda \in [0,1]$ and we
predict an error at position $i$ in sequence $\omega$ if and only if the probability $P(i,n)$ of a nucleotide $n \in \{ A,C,G,U \}$ exceed this threshold.
The set of corrections is thus $\{ n \; | \;  n \in \{ A,C,G,U \} \mbox{ and } P(i,n) > \lambda \mbox{ and }  n \neq \omega[i] \}$, where $\omega[i]$ is the
nucleotide at position $i$ in the input sequence. Then, we progressively vary $\lambda$ between $0$ and $1$ to calculate the ROC curve and the AUC.

\begin{table}
\begin{center}
\begin{tabular}{|c|c|c|c|}
\hline
& 1.0 & 0.5 & 0. \\
\hline
6 & 0.69 & 0.72 & 0.74 \\
12 & 0.80 & 0.84 & 0.84 \\
24 & 0.76 & 0.84 & 0.84 \\
\hline
\end{tabular}
\end{center}
\caption{Performance of error-correction. The row index indicates the number of mutations performed with \RNApyro, and the column index indicates
the value of the parameter $\alpha$ distributing the weights of stacking pair energies vs isostericity scores. In each cell, we report the average AUC
values over the 10 experiments.}
\label{tab:benchmark}
\end{table}

\begin{figure}
\begin{subfigure}[b]{0.3\textwidth}
\includegraphics[width=1.2\textwidth]{figures/ROC_6.pdf}
\caption{6 mutations}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.3\textwidth}
\includegraphics[width=1.2\textwidth]{figures/ROC_12.pdf}
\caption{12 mutations}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.3\textwidth}
\includegraphics[width=1.2\textwidth]{figures/ROC_24.pdf}
\caption{24 mutations}
\end{subfigure}
\end{figure}

\subsection{Time}
The applications for this algorithms implies a need for efficiency
and scalability. In Table~\ref{tab:time} we
present different times needed to compute the probabilities for
 every nucleotide at every positions for a vast set of parameters. For those test
 all the sequences were generated randomly, as the target secondary structure.

>>>>

We build our data set from the 5S rRNA multiple sequence alignment (MSA) available in the Rfam Database 11.0 (Rfam id: \texttt{RF00001}).
Since our software does not currently implement gaps (mainly because scoring indels is a challenging issue that cannot be fully addressed
in this work),  we clustered together the sequences with identical gap locations. From the $54$ MSAs without gap produced, we selected the
biggest MSA  which contains $130$ sequences (out of $712$ in the original Rfam MSA). Then, in order to avoid any bias, we used \texttt{cd-hit}
\cite{Li:2006fk} to remove sequences with more than 80\% of sequence similarity. This operation resulted in a data set of $45$ sequences.

We design our benchmark using a leave-one-out strategy. We randomly picked one sequence from our data set and performed $12$ random
mutations. Our sequences have $119$ nucleotides, thus the number of mutations corresponds to an error-rate of 10\%. We repeated this operation 
$10$ times. The value of $\beta$ has been set to $1.5$ (larger values gave similar results). To estimate the impact of the distribution of the weights
between the energy term and isostericity  score, we use 4 different values of $\alpha = {0, 0.5, 0.8, 1.0}$. Similarly, we also investigate the impact of 
an under- and over- estimate of the number of errors, and use values equal to 50\% (6 mutations) and 200\% (24 mutations) of the exact number of
errors (i.e. 12).

To evaluate our method, we computed a ROC curve representing the performance of a classifier based on the mutational probabilities computed
by \RNApyro. More specifically, we fix a threshold $\lambda \in [0,1]$ and we predict an error at position $i$ in sequence $\omega$ if and only if the
probability $P(i,n)$ of a nucleotide $n \in \{ A,C,G,U \}$ exceed this threshold. To correct the errors we use the set of nucleotides with a probability
higher than this threshold is  $C(i) = \{ n \; | \;  n \in \{ A,C,G,U \} \mbox{ and } P(i,n) > \lambda \mbox{ and }  n \neq \omega[i] \}$, where $\omega[i]$ is
the nucleotide at position $i$ in the input sequence. We note that for the lowest thresholds multiple nucleotides can be available in $C(i)$ to correct
the sequence. Here, we remind that we aim to estimate the potential of error-correction of \RNApyro and not to develop a error-correction pipe-line. 
The latter will be investigate in further studies. Finally, we progressively vary $\lambda$ between $0$ and $1$ to calculate the ROC curve and the area
under the curve (AUC). We report our results in Figure \ref{fig:ROCall}.

\begin{figure}
\begin{subfigure}[b]{0.3\textwidth}
\includegraphics[width=1.2\textwidth]{figures/ROC_6.pdf}
\caption{6 mutations}
\label{fig:ROC6mut}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.3\textwidth}
\includegraphics[width=1.2\textwidth]{figures/ROC_12.pdf}
\caption{12 mutations}
\label{fig:ROC12mut}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.3\textwidth}
\includegraphics[width=1.2\textwidth]{figures/ROC_24.pdf}
\caption{24 mutations}
\label{fig:ROC24mut}
\end{subfigure}
\caption{Performance of error-correction. The row index indicates the number of mutations performed with \RNApyro, and the column index indicates
the value of the parameter $\alpha$ distributing the weights of stacking pair energies vs isostericity scores. In each cell, we report the average AUC
values over the 10 experiments.}
\label{fig:ROCall}
\end{figure}

Our data demonstrate that our algorithm exhibits interesting performance for error-correction applications. First, the AUC values (up to $0.86$) indicate that a
signal has been successfully extracted. This result has been achieved with errors in loop regions (i.e. without base pairing information) and thus suggests
that correction rates in structured regions (i.e. base paired regions) could be even higher. Next, the optimal values of $\alpha$ tend to be close to $0.0$. This 
finding suggests that at this point the information issued from the stacking energy is currently modest. However, specific examples showed improved performance
using this energy term. Further studies must be conducted to understand how to make the best use it. Finally, our algorithm seems robust to the number of
mutations considered. Indeed, good AUC values are achieved with low estimates of the number of errors in the sequence (c.f. 50\% of the errors  in
Fig.~\ref{fig:ROC6mut}) and with large  values (c.f.~200\% of the errors  in Fig.~\ref{fig:ROC24mut}). It is worth noting that scoring scheme with larger weight on
the isostericity scores (low $\alpha$ values) seem more robust to under- and over-estimate of the number of errors.

\subsection{Time}
Further NGS applications of \RNApyro implies a need for efficiency and scalability. In Table~\ref{tab:time} we
present different times needed to compute the probabilities for  every nucleotide at every positions for a vast set of parameters. For those test
 all the sequences were generated randomly, as the target secondary structure.
